\documentclass[11pt,onecolumn]{article}
\usepackage{fancyvrb}
\usepackage{relsize}
\usepackage{fullpage}
\usepackage{amsmath} % for \text; HeVeA doesn't know about amstext
\usepackage{latexsym} % for \Diamond
\usepackage{hevea}
\usepackage{comment}
\usepackage[dvips]{graphicx}       %%% graphics for dvips
\usepackage[colorlinks=true, linkcolor=blue, 
  citecolor=blue, urlcolor=blue,
  ps2pdf,                %%% hyper-references for ps2pdf
  bookmarks=true,        %%% generate bookmarks ...
  bookmarksnumbered=true,%%% ... with numbers
]{hyperref}
% pdfcreator and pdfproducer are set automatically in pdfLaTeX
\hypersetup{ pdfcreator  = {LaTeX with hyperref package},
  pdfproducer = {dvips + ps2pdf} }
\begin{latexonly}
% HeVeA can't handle this:
\let\url\nolinkurl % because dvips cannot break url across lines
\end{latexonly}
%\usepackage{times}

%\includecomment{svn}
%\excludecomment{release}
\includecomment{release}
\excludecomment{svn}

\begin{svn}
\newcommand{\sreither}[2]{#1}
\end{svn}
\begin{release}
\newcommand{\sreither}[2]{#2}
\end{release}

% Make \_ be CMTT's underscore, not \textunderscore, since we use
% it in code and command names.
\def\_{\char"5F}
\newcommand{\todo}[1]{\textbf{[[#1]]}}
\newcommand{\mytt}{\small \tt}

\newcommand{\titled}{Vine installation and user manual}
\title{\mbox{}\\[-.8in]\bf \titled}
\author{BitBlaze Team}
\date{August 26th, 2009:
\sreither{SVN trunk r4183}{Release 1.0}
and Ubuntu 9.04}
\begin{document}
\maketitle

\tableofcontents

\section{Introduction}

This document is a guide for setting up and running Vine, the static
analysis component of the BitBlaze Binary Analysis Framework. It
assumes that you have some familiarity with Linux.  The instructions
are based on
\begin{svn}
the version of Vine in the SVN trunk as of the date shown
in the header,
\end{svn}
\begin{release}
the release of Vine mentioned in the header,
\end{release}
running on a vanilla Ubuntu 9.04 distribution of Linux.
It includes details about how to install Vine, and then walks through
a simple usage example intermixed with explanations about the tools
used.

The example takes a trace from a simple program with symbolic keyboard
input, and generates an STP file which models the weakest precondition
of the control-flow path the program took. In other words, the
conditions on the inputs that cause it to take a execute a certain
path of code. The trace file was generated using TEMU, the dynamic
component of the BitBlaze Binary Analysis Framework, but it is
included in the {\tt examples} directory of the Vine distribution so
you can try out the tool without using TEMU.

\section {\label{sec:install}Installation}

Vine is written using a combination of OCaml and C++, and distributed
in source code form.
%
Therefore the main steps in installing it are installing the
prerequisite software it requires, and then compiling the source code.
%
To install prerequisite software, we recommend that you use a recent
version of Ubuntu Linux, for which we have verified that all the
needed packages are already available.
%
The compilation is performed automatically using {\tt configure} and
{\tt Makefile} scripts like many other Linux applications.
%
The following subsections cover these tasks in more detail, and then
we finish by giving complete listing of the commands needed for our
recommended platform.

\subsection{\label{sec:prereqs}Prerequisites}

Our recommended platform for using Vine is a 32-bit x86 version of
Ubuntu Linux, version 9.04 (code named ``Jaunty Jackalope''); we used
such a system in preparing these instructions.
%
The needed packages all also exist in Debian Linux, so the process
there should work in almost the same way.
%
It is possible to use Vine with other Linux distributions, but you may
need to compile some of the prerequisite software from source.
%
The pure OCaml parts of Vine also work fine on 64-bit x86-64 Linux
platforms, but the library it uses for the semantics of x86
instructions expects to run on a 32-bit platform when processing
32-bit code, so a 32-bit platform is needed to make complete use of
Vine.
%
If you are using an x86-64 version of Ubuntu or Debian, we recommend
installing a parallel 32-bit version of your OS packages using a
mechanism called a ``chroot'', but how to do so is beyond the scope of
this manual (or you could use another kind of virtual machine).

If you don't have any of the prerequisites already installed, they
will require about 300MB to download, and take up about 1.1GB of disk
space once installed.
%
A build of Vine itself requires about 320MB.

\begin{svn}
\subsubsection{Subversion}

Both the development versions of Vine itself and the VEX library we
use are maintained using the Subversion version control system, so
you'll need it to download them. In Ubuntu 9.04, version 1.5.4 is
available in the package {\tt subversion}.
\end{svn}

\begin{svn}
\subsubsection{Autoconf and Automake}

If you are building from the SVN development version of Vine, you will
need to regenerate the {\tt configure} and {\tt Makefile.in} files
needed by the compile using the automake and autoconf tools. In Ubuntu
9.04, you can get suitable versions of both (Automake 1.10.2 and
Autoconf 2.63) by requesting the {\tt automake} package.
\end{svn}

\subsubsection{C++ compiler}

For compiling the C++ code in Vine, we recommend the G++ compiler from
GCC which is standard on Linux; the default version in Ubuntu 9.04 is
4.3.3, and the package name is {\tt g++}.

\subsubsection{\label{sec:vex}The VEX library}

Vine uses the VEX library (which is also used by Valgrind) to get
information about the behavior of x86 instructions.
%
The current version of Vine is designed to work with SVN revision
r1856 of VEX, which is maintained in the Valgrind SVN repository at
\url{svn://svn.valgrind.org/vex/trunk}.
%
\begin{svn}%
You can check it out into a directory named {\tt vex} using the
following command:

\begin{Verbatim}[frame=lines, framesep=.5em]
svn checkout -r1856 svn://svn.valgrind.org/vex/trunk vex
\end{Verbatim}

Then, to compile it, run the following commands in the {\tt vex}
directory:

\begin{Verbatim}[frame=lines, framesep=.5em]
make version
make libvex_x86_linux.a
make libvex.a
\end{Verbatim}
\end{svn}
\begin{release}
For your convenience, we have included an appropriate version of VEX
with the release, and it will be compiled automatically.
\end{release}

\subsubsection{STP}

Vine interfaces with STP, a satisfiability-modulo-theories (SMT)
decision procedure for bit vector (bounded integer) arithmetic and
arrays.
%
Vine can either interact directly with the prover through a
programmatic interface, or it can produce formulas in STP's text
format.
%
The directory {\tt stp} contains x86 and x86-64 binaries for a version
of STP we have tested to work well with Vine.
%
If you would like to use a different version of STP, it is available
in source form at
\url{http://people.csail.mit.edu/vganesh/STP_files/stp.html}.

\subsubsection{Binutils libraries}

Vine uses the GNU BFD (Binary File Descriptor) library and related
libraries from the GNU Binutils to parse the structure of executables,
and for human-readable dissassembly. On Ubuntu they are distributed in
a package named {\tt binutils-dev} (version 2.19.1 in Ubuntu 9.04).

(A technical legal point: the VEX library with which Vine links is
distributed under version 2 of the GNU GPL only, whereas versions 2.18
and later of the GNU Binutils are distributed only under versions 3
and later of the GNU GPL.
%
Unfortunately, versions 2 and 3 of the GPL are mutually incompatible,
so if you plan to distribute copies of Vine for platforms where the
Binutils are not system libraries in the sense of the GPL, you may
wish to use version 2.17 or earlier of the Binutils instead.)

\subsubsection{OCaml build tools}

Most of Vine is implemented in the functional language OCaml, so OCaml
development tools are required.
%
In addition to the standard OCaml compiler and tools, Vine uses the
Findlib library for managing OCaml packages, and the CamlIDL tool for
generating interfaces to C code.
%
In Ubuntu 9.04, these are available as the {\tt ocaml} package
(version 3.10.2), the {\tt ocaml-findlib} package (version 1.2.1), and
the {\tt camlidl} package (version 1.05).
%
Optionally, you can also install the natively compiled versions of the
OCaml build tools (which are somewhat faster) from the
{\tt ocaml-native-compiler} package.

\subsubsection{\label{sec:ocamlgraph}ocamlgraph}

The ocamlgraph library provides graph data structures and algorithms,
which Vine uses to represent control flow graphs.
%
Unfortunately, an interface that Vine uses changed incompatibly in
version 0.99 of the library, so the source code we distribute supports
only version 0.99 and later.
%
In Ubuntu 9.04, it is available as the package {\tt
libocamlgraph-ocaml-dev}.
%
Note, however, that the version of the library packaged in earlier
versions of Ubuntu is not compatible.
%
If a recent version of ocamlgraph is not packaged for your system, you
have several options:
\begin{enumerate}
\item {\bf Compile ocamlgraph from source.}
You can obtain the source for the latest version of ocamlgraph from
\url{http://ocamlgraph.lri.fr/}.
%
Note that it only supports installation in {\tt /usr/lib} or {\tt
/usr/local/lib}.
%
Also, some versions of ocamlgraph have a Makefile bug that causes them
to look for interface files in the wrong location, which can be fixed
by applying the following patch to {\tt Makefile.in}:

\begin{Verbatim}[frame=lines, framesep=.5em]
--- Makefile.in.orig
+++ Makefile.in
@@ -210,7 +210,7 @@
 
 install-findlib: META
 ifdef OCAMLFIND
-	$(OCAMLFIND) install ocamlgraph META *.mli \
+	$(OCAMLFIND) install ocamlgraph META $(LIBDIR)/*.mli \
 		graph$(LIBEXT) graph.cmx graph.cmo graph.cmi $(CMA) $(CMXA)
 endif
\end{Verbatim}

\item {\bf Make a backport package.}
If you would like to have the installation of ocamlgraph managed by
your regular package manager, another option is to build a package
yourself. For instance, a suitable package for Ubunutu 8.04 can be
built and installed using Debian sources, as follows:

\begin{Verbatim}[frame=lines, framesep=.5em]
sudo apt-get install   libocamlgraph-ocaml-dev
sudo apt-get build-dep libocamlgraph-ocaml-dev
sudo apt-get install liblablgtk2-ocaml-dev liblablgtk2-gnome-ocaml-dev \
                     docbook-xsl po4a
sudo apt-get install fakeroot
svn co svn://svn.debian.org/svn/pkg-ocaml-maint/trunk/packages/ocamlgraph \
       -r5983
tar xvzf ocamlgraph/upstream/ocamlgraph_0.99c.orig.tar.gz
mv ocamlgraph/trunk/debian ocamlgraph-0.99c
perl -pi -e 's[ocaml-nox \(>= 3.10.0-9\)] #\
              [ocaml-nox  (>= 3.10.0-8)]' ocamlgraph-0.99c/debian/control
(cd ocamlgraph-0.99c && dpkg-buildpackage -us -uc -rfakeroot)
sudo dpkg -i libocamlgraph-ocaml-dev_0.99c-2_i386.deb
\end{Verbatim}

\item {\bf Patch the Vine source.}
Because Vine does not use the extra functionality introduced in the
new library version, another option is to change the Vine code that
uses the library back to the older interface.
%
This requires four one-line changes in four files under {\tt ocaml},
as in the following patch:

\begin{Verbatim}[frame=lines, framesep=.5em]
Index: ocaml/vine_cfg.mli
===================================================================
--- ocaml/vine_cfg.mli
+++ ocaml/vine_cfg.mli
@@ -299,7 +299,7 @@
 
 module Component :
 sig
-  val scc : G.t -> int * ( G.V.t -> int )
+  val scc : G.t -> G.V.t -> int
   val scc_list : G.t -> G.V.t list list
 end
 
Index: ocaml/vine_cfg.ml
===================================================================
--- ocaml/vine_cfg.ml
+++ ocaml/vine_cfg.ml
@@ -1271,7 +1271,7 @@
     if cfg#has_edge rb ra then false
     else (cfg#add_edge rb ra; true) (* temporary backedge *)
   in
-  let (_,scc) = Component.scc cfg in
+  let scc = Component.scc cfg in
   let group = scc a in
   let newcfg = new cfg 8 cfg#get_iter_labels_function cfg#vardecls in
   let (outside: 'a bb) =
Index: ocaml/vine_callstring.ml
===================================================================
--- ocaml/vine_callstring.ml
+++ ocaml/vine_callstring.ml
@@ -178,7 +178,7 @@
   else 
     csg.graph 
   in 
-  let (_,scc) = Component.scc g in 
+  let scc = Component.scc g in 
   let group = scc a in 
   let addifgroup v newg = 
     if scc v = group then
Index: utils/chop.ml
===================================================================
--- utils/chop.ml
+++ utils/chop.ml
@@ -243,7 +243,7 @@
       fun () -> cfg#remove_edge t s 
     ) 
   in
-  let (_,scc) = Component.scc cfg in
+  let scc = Component.scc cfg in
     (* make sure we really have a cycle *)
   let () = assert(scc (cfg#get_id s) = scc (cfg#get_id t)) in 
   let group = scc (cfg#get_id s) in
\end{Verbatim}
\end{enumerate}

\subsubsection{Other OCaml libraries}

Vine also uses several further OCaml libraries:
\begin{itemize}
\item ExtLib provides an extended standard library (e.g., more data
  structures) for OCaml; version 1.5.1 is in the package {\tt
    libextlib-ocaml-dev}.
\item GDome2 is a document object model for dealing with XML documents
  that Vine uses via its OCaml bindings. Version 0.2.6 is available as
  the package {\tt libgdome2-ocaml-dev}.
\end{itemize}

\begin{svn}
\subsubsection{SQLite and its OCaml bindings}

Vine uses the lightweight (serverless) SQL library SQLite, via an
OCaml binding, to store databases of disassembled code with its IDA Pro
static disassembler plug-in.
%
On Ubuntu 9.04, you should install the library itself (version 3.6.10)
via the package {\tt libsqlite3-0} and the OCaml interface via the
{\tt libsqlite3-ocaml-dev} package.
%
It is also convenient to install the command-line database tool, in
the {\tt sqlite3} package.
\end{svn}

\subsubsection{\LaTeX}

Vine's documentation (including this document) is written using the
\LaTeX\ markup language, so you will need to install it to rebuild the
documentation.
%
On Ubuntu 9.04, the needed parts are included under the {\tt texlive},
{\tt texlive-latex-extra}, and {\tt transfig} packages.
%
To build an HTML version of the documentation, we also use \hevea,
which is in the {\tt hevea} package.

\subsection{Compiling}

\begin{svn}
\subsubsection{SVN checkout}

The primary development version of Vine is referred to as the
``trunk'' in the context of our Subversion repository, and it's kept
at the URL \url{https://bullseye.cs.berkeley.edu/svn/vine/trunk}.
%
Note that this repository is password-protected; if someone has
pointed you at this version of these instructions, you should already
have an appropriate account; if you don't, you'll need to get one.
%
For instance, you can check out a copy of the source code into a
directory named {\tt vine} by using the command:

\begin{Verbatim}[frame=lines, framesep=.5em]
svn checkout https://bullseye.cs.berkeley.edu/svn/vine/trunk vine
\end{Verbatim}

\subsubsection{Autoconf and Automake}

To avoid redundancy in the repository, it doesn't store a copy of the
{\tt configure} script or related files, so you will need to
regenerate them by running the {\tt autogen.sh} script; it doesn't
require any arguments.
\end{svn}

\begin{release}
\subsubsection{Unpacking}

Vine is distributed as a gzip-compressed tar archive, which you can
unpack into a directory {\tt vine-1.0} using the command ``{\tt tar xvzf
vine-1.0.tar.gz}''.

\end{release}

\subsubsection{Configure}

To prepare the Vine source for compilation, you'll need to run the
{\tt configure} script in the Vine source directory.
%
The script accepts all of the standard options and environment
variables for autoconf-based configure scripts, though most should not
be necessary.
\begin{svn}
One important option is {\tt --with-vex=}, which you should provide
with the path to the VEX source directory you compiled as described in
section~\ref{sec:vex}.
\end{svn}

\subsubsection{Build code}

After the configuration script has finished, you can compile Vine by
running {\tt make} in the top-level Vine directory.
%
This will compile first the C++ library and then the OCaml modules.

\subsubsection{Build documentation}

To generate the documentation that comes with Vine, go to the {\tt
  vine/doc} subdirectory and give the command ``{\tt make doc}''.

\subsection{Summary}

To recap the steps described above, we now show a script for all of
the commands needed to compile Vine, starting with a fresh
installation of Ubuntu 9.04.
\begin{svn}
(This is also found as the file
\verb'docs/install-vine-svn.sh' in the Vine source.)

\VerbatimInput{../install-vine-svn.sh}
\end{svn}
\begin{release}
(This is also found as the file
\verb'docs/install-vine-release.sh' in the Vine source.)

\VerbatimInput{../install-vine-release.sh}
\end{release}

\input{vine-overview}

\input{vine-il}

\section{Example}

We now illustrate the use of Vine with an example.
%
In it, we will take a trace generated by TEMU from a program that
parses an integer and checks whether it is equal to 5.
%
We will use Vine to build a version of the execution path in which the
input is symbolic, and compute a path condition: a formula over the
inputs which, if true, causes execution to take the same path.
%
Finally, we will use STP to solve the path condition and reconstruct
an input that would cause the program to take the same path.
%
The trace is included in the {\tt examples} directory under the name
{\tt five.trace}.

\input {irformulagen}
\input {stp}

\input {utils}

\section{Troubleshooting}

This section lists some errors that you may encounter while using
Vine, and gives suggestions on resolving them.

\begin{itemize}
\item Incompatible types in Vine\_cfg

\begin{Verbatim}[frame=lines, framesep=.5em]
File "vine_cfg.ml", line 1301, characters 16-33:
This expression has type G.V.t -> int but is here used with type 'a * 'b
\end{Verbatim}

This error occurs if you try to compile Vine with a version of the
ocamlgraph library older than 0.99, which has an incompatible type for
one function.
It can be avoided by using a newer version of the library, or worked
around by modifying the Vine source; see Section~\ref{sec:ocamlgraph}
for more details.

\item Size assertion in VEX

\begin{Verbatim}[frame=lines, framesep=.5em]
vex: priv/host-x86/hdefs.c:2332 (emit_X86Instr):
Assertion `sizeof(UInt) == sizeof(void*)' failed.
\end{Verbatim}

This error occurs if you try to use a 64-bit version of Vine to
process 32-bit x86 code.
Because the VEX library does not support cross-platform operation,
Vine can only translate x86 code when compiled in 32-bit mode.
However, you can still compile and run an x86 version of Vine on an
x86-64 platform (see Section~\ref{sec:prereqs} for further
discussion).
You can also generate a Vine IL file on a 32-bit platform and then do
further processing on a 64-bit one.

\item OCaml stack overflow

\begin{Verbatim}[frame=lines, framesep=.5em]
Fatal error: exception Stack_overflow
\end{Verbatim}

This error occurs when an OCaml program tries to use more stack space
than is available.
If it occurs even on a very small input, it could be caused by an
infinite recursion bug, but more commonly it is caused by processing a
large data structure with a recursive algorithm.
One potential fix is to increase the amount of stack space available.
For native-compiled OCaml programs, stack usage is limited by the
operating system's stack size resource limit, which may have a small
default value such as 8MB.
You can remove this limit with a shell command, such as {\tt ulimit -s
  unlimited} in an sh-style shell or {\tt limit stacksize unlimited}
in a csh-style shell; see your shell's documentation for more details.
Sometimes debugging versions of programs use more stack space, so if
you encounter this error with the {\tt .dbg} version of a program, try
the version without that suffix.
If the error was caused by recursion, a stack backtrace should reveal
what function was the culprit; to obtain one, rerun the program with
the {\tt OCAMLRUNPARAM} environment variable set to {\tt b}.
\end{itemize}

\section {Reporting Bugs}

\begin{svn}
Please report bugs to the bugzilla at:
\texttt{https://bullseye.cs.berkeley.edu/bugzilla/}.

When reporting bugs in the SVN version of the software, it's most
useful if you can reproduce your problem with the most recent trunk
version, but if you're using an older version, please specify the
revision number (i.e., the output of the svnversion command in your
bug report). And please also report if you notice something wrong or
out of date in this document.
\end{svn}
\begin{release}
Though we cannot give any guarantee of support for Vine, we are
interested in hearing what you are using it for, and if you encounter
any bugs or unclear points.
%
Please send your questions, feature suggestions, bugs (and, if you
have them, patches) to the bitblaze-users mailing list.
%
Its web page is:
\url{http://groups.google.com/group/bitblaze-users}.
\end{release}
\end{document}
